---
title: "dplyr_data_transformation"
output: html_document
date: "2025-06-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Visualization is an important tool for generating insight, but it’s rare that you get the data in exactly the right form you need to make the graph you want. Often you’ll need to create some new variables or summaries to answer your questions with your data, or maybe you just want to rename the variables or reorder the observations to make the data a little easier to work with. You’ll learn how to do all that (and more!) in this chapter, which will introduce you to data transformation using the dplyr package

```{r}
library(nycflights13)
library(tidyverse)

flights
```

The most important difference between tibbles and data frames is the way tibbles print; they are designed for large datasets, so they only show the first few rows and only the columns that fit on one screen. There are a few options to see everything.
```{r}
glimpse(flights)
```
Format of dplyr verbs:

The first argument is always a data frame.

The subsequent arguments typically describe which columns to operate on using the variable names (without quotes).

The output is always a new data frame.

```{r}
flights |>
  filter(dest == "IAH") |>
  group_by(year,month,day) |>
  summarize(
    arr_delay = mean(arr_delay, na.rm=TRUE)
  )
```

dplyr’s verbs are organized into four groups based on what they operate on: rows, columns, groups, or tables. 

How to work on rows:
filter(), which changes which rows are present without changing their order, and arrange(), which changes the order of the rows without changing which are present.

```{r}
flights |>
  filter(dep_delay > 120) ## > , <, = >=, <=, !=, &, |
```
```{r}
flights |>
  filter(month == 1 & day == 1)

```
```{r}
flights |>
  filter(month == 1 | month == 2)
```
Matching at least one of the values
```{r}
jan_feb = flights |>
  filter(month %in% c(1,2))
```

arrange()

arrange() changes the order of the rows based on the value of the columns. It takes a data frame and a set of column names (or more complicated expressions) to order by. If you provide more than one column name, each additional column will be used to break ties in the values of the preceding columns.

```{r}
flights |>
  arrange(year, month, day, dep_time)
```
We can use desc() on a column inside of arrange() to re-order the data frame based on that column in descending (big-to-small) order

```{r}
flights |>
  arrange(desc(dep_delay))
```
distinct() finds all the unique rows in a dataset, so technically, it primarily operates on the rows.

```{r}
flights |>
  distinct()

flights |>
  distinct(origin, dest)

flights |>
  distinct(origin, dest, .keep_all = TRUE)

```
To count number of occurences
```{r}
flights |>
  count(origin, dest, sort = TRUE)
```
How to work on columns

There are four important verbs that affect the columns without changing the rows: mutate() creates new columns that are derived from the existing columns, select() changes which columns are present, rename() changes the names of the columns, and relocate() changes the positions of the columns.


mutate

The job of mutate() is to add new columns that are calculated from the existing columns. 
```{r}
flights |>
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time *60
  )



```
By default, mutate() adds new columns on the right-hand side of your dataset, which makes it difficult to see what’s happening here. We can use the .before argument to instead add the variables to the left-hand side

```{r}
flights |>
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time *60,
    .before=1
  )

```
The . indicates that .before is an argument to the function, not the name of a third new variable we are creating. You can also use .after to add after a variable, and in both .before and .after you can use the variable name instead of a position. 

```{r}
flights |>
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time *60,
    .after = day
  )
```

Now, is select. 
It allows you to rapidly zoom in on a useful subset using operations based on the names of the variables.
Select columns by name
```{r}
flights |> 
  select(year, month, day)
```
Select all columns between year and day
```{r}
flights |> 
  select(year:day)
```

Select all columns except those from year to day (inclusive)

```{r}
flights |>
  select(!year:day)
```

Select all columns that are characters:

```{r}
flights |> 
  select(where(is.character))
```
You can rename variables as you select() them by using =. The new name appears on the left-hand side of the =, and the old variable appears on the right-hand side

```{r}
flights |> 
  select(tail_num = tailnum)
```

rename() - If you want to keep all the existing variables and just want to rename a few, you can use rename() instead of select():

```{r}
flights |> 
  rename(tail_num = tailnum)
```

relocate()  - Use relocate() to move variables around. You might want to collect related variables together or move important variables to the front. By default relocate() moves variables to the front


```{r}
flights |> 
  relocate(time_hour, air_time)
```
You can also specify where to put them using the .before and .after arguments, just like in mutate():

```{r}
flights |> 
  relocate(year:dep_time, .after = time_hour)
flights |> 
  relocate(starts_with("arr"), .before = dep_time)
```
The pipe - its real power arises when you start to combine multiple verbs

```{r}
flights |> 
  filter(dest == "IAH") |> 
  mutate(speed = distance / air_time * 60) |> 
  select(year:day, dep_time, carrier, flight, speed) |> 
  arrange(desc(speed))
```
Instead of pipe, following also works. But pipe is easier to understand.
```{r}
arrange(
  select(
    mutate(
      filter(
        flights, 
        dest == "IAH"
      ),
      speed = distance / air_time * 60
    ),
    year:day, dep_time, carrier, flight, speed
  ),
  desc(speed)
)
```

Following also works but too many lines of code
```{r}
flights1 <- filter(flights, dest == "IAH")
flights2 <- mutate(flights1, speed = distance / air_time * 60)
flights3 <- select(flights2, year:day, dep_time, carrier, flight, speed)
arrange(flights3, desc(speed))
```
Groups

dplyr gets even more powerful when you add in the ability to work with groups. In this section, we’ll focus on the most important functions: group_by(), summarize(), and the slice family of functions.

group_by() - Use group_by() to divide your dataset into groups meaningful for your analysis

```{r}
flights |> 
  group_by(month)
```
summarize() - The most important grouped operation is a summary, which, if being used to calculate a single summary statistic, reduces the data frame to have a single row for each group.

```{r}
flights |> 
  group_by(month) |> 
  summarize(
    avg_delay = mean(arr_delay, na.rm=TRUE) ## avoid NA values
  )
```
You can create any number of summaries in a single call to summarize(). You’ll learn various useful summaries in the upcoming chapters, but one very useful summary is n(), which returns the number of rows in each group:


```{r}
flights |> 
  group_by(month) |> 
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE), 
    n = n()
  )
```
The slice_ functions
There are five handy functions that allow you to extract specific rows within each group:
df |> slice_head(n = 1) takes the first row from each group.
df |> slice_tail(n = 1) takes the last row in each group.
df |> slice_min(x, n = 1) takes the row with the smallest value of column x.
df |> slice_max(x, n = 1) takes the row with the largest value of column x.
df |> slice_sample(n = 1) takes one random row.
```{r}
flights |> 
  group_by(dest) |> 
  slice_max(arr_delay, n = 1) |>
  relocate(dest)
```
Grouping by multiple variables

```{r}
daily <- flights |>  
  group_by(year, month, day)

daily_flights <- daily |> 
  summarize(n = n())
#> `summarise()` has grouped output by 'year', 'month'. You can override using
#> the `.groups` argument.
```
If you’re happy with this behavior, you can explicitly request it in order to suppress the message:


```{r}
daily_flights <- daily |> 
  summarize(
    n = n(), 
    .groups = "drop_last"
  )
```

Ungrouping - You might also want to remove grouping from a data frame without using summarize(). You can do this with ungroup().

```{r}
daily |> 
  ungroup()
```
Now let’s see what happens when you summarize an ungrouped data frame.

```{r}
daily |> 
  ungroup() |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE), 
    flights = n()
  )
```

.by - per-operation grouping, the .by argument

```{r}
flights |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE), 
    n = n(),
    .by = month
  )
```
Or if you want to group by multiple variables:
```{r}
flights |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE), 
    n = n(),
    .by = c(origin, dest)
  )
```


